In-silico molecular enrichment and clearance of the human intracranial space

### 2025-02-25 paper/main.tex
Updates from Overleaf

\subsection*{In-silico predictions of intracranial molecular enrichment and clearance after intrathecal injection}
\begin{figure}
    \centering
    \includegraphics[width=0.93\textwidth]{figures/figure1.png}
     \caption{
     \textbf{Our in-silico model predicts tracer transport in PVS, SAS and parenchyma following intrathecal injection.}\\
     A) Illustration of the spatial domains of the model: CSF-filled space (ventricles and SAS) $\Omega_{\rm CSF}$, parenchyma $\Omega_{\rm PAR}$ and the PVS surrounding arteries $\Lambda_{\rm artery}$ and the veins $\Lambda_{\rm vein}$ and their interfaces and boundaries (see Methods); B) 3D rendering of the actual geometry used in this study parenchyma, CSF (clipped, blue), arterial network (red), venous network (blue)); C) arterial network (red) and venous network (blue); D) illustration of the modelled clinical tracer injection procedure; E) time of simulated first tracer arrival after intrathecal injection (concentration exceeds $0.1\,$mol/l); F) time of first tracer arrival in the arterial PVS (top) and venous PVS (bottom); G) sagittal, coronal and axial view of computed tracer concentrations over T1-image 1, 6, 12 and 24\,h after injection (low concentrations transparent, pial surface in pink, arachnoid membrane in cyan, and arteries in dark red); H) total amount of tracer in each cranial compartment over the first 24\,h; I) average tracer concentration in each compartment over the first 24\,h. }
     \label{fig:results1}
\end{figure}
Using previously published multi-modal magnetic resonance imaging
(MRI) data~\cite{hodneland2019new, deistung2009tof,
  schweser2012quantitative, reichenbach2012future,
  deistung2017overview}, we construct a multiscale computational
representation of the human intracranial compartments consisting of
the CSF spaces and brain parenchyma as three-dimensional (3D) domains
and with the PVSs surrounding major surface arteries and veins as
embedded networks of topologically one-dimensional (1D) curves
(\Cref{fig:results1}A--C). We consider a solute concentration field,
varying in space and time, in the 3D domains and in the PVS networks,
and assume that the solute can cross between these compartments
through semi-permeable membranes. As the drivers and modes of
intracranial transport are under substantial
debate\cite{smith2019going, proulx2021cerebrospinal,
  bohr2022glymphatic, hladky2022glymphatic, betsholtz2024advances},
our first target is to establish a baseline model accounting for a
reasonably conservative set of mechanisms and their integrated effect
over a timescale of several minutes to a few days. To this end, we
assume that the solute will (i) diffuse within all compartments, with
diffusivity depending on the effective properties of the relevant
medium\cite{sykova2008diffusion}; (ii) experience significant
dispersive effects due to the pulsatile flow of CSF induced by the
cardiac and respiratory cycles\cite{vinje2019respiratory,
  sharp2019dispersion, ray2021quantitative, troyetsky2021dispersion};
and (iii) be convected by a (small) net flow of CSF resulting from
production in the choroid plexus with CSF efflux across the upper
convexity\cite{hornkjol2022csf} and from the peristaltic pumping
effect of pulse wave pulsations in surface periarterial
spaces~\cite{mestre2018flow, gjerde2023directional}. Mathematically, this model is represented by a
mixed-dimensional system of coupled time-dependent partial
differential equations\cite{masri2024modelling}, which we solve
numerically with high accuracy using a mass-conserving finite element
scheme and the FEniCS finite element software\cite{alnaes2015fenics,
  kuchta2020assembly} (see Methods). The computational framework and
associated software are all openly available~\cite{ZENODO}.

Simulating a glymphatic MRI protocol~\cite{ringstad2017glymphatic,
  ringstad2018brain, eide2024functional}(\Cref{fig:results1}D), we then predict the
spreading of $0.5$ mmol intrathecally injected Gadobutrol after its
appearance at the craniocervical junction
(\Cref{fig:results1}E--G). After one hour, the in-silico
tracer moves upwards in the SAS frontally of the brainstem, and
quickly reaches the supratentorial regions. Here, it spreads both
posterior through the quadrigeminal cistern and the longitudinal
fissure, and anteriorly through the outer SAS, reaching the top of the
cerebral cortex after around 12 hours. After 24 hours, the tracer
covers most of the brain surface (with the exception of some posterior
regions) and has penetrated substantially into the parenchymal
tissue. This pattern is reflected in the mean tracer concentrations in
each compartment, where the CSF space reaches its peak of
$1.4\,$mmol/l after 2 hours, followed by the arterial PVS
concentration peak ($2.9\,$mmol/l after 4 hours)
(\Cref{fig:results1}I). While the mean concentrations in these
compartments drop soon after peaking, the parenchymal tissue
slowly enriches with tracer over the first 24 hours, with a final
value of $0.3$ mmol/l. Overall, about 40\% of the total amount of
tracer remains in the cranium 24 hours post-injection, with the largest
share in the CSF (53\%), followed by the parenchyma (35\%), the
venous PVS (8\%) and the arterial PVS
(5\%) (\Cref{fig:results1}H).

>>>>

\subsection*{In-silico predictions of intracranial molecular enrichment and clearance after intrathecal injection}
\begin{figure}
    \centering
    \includegraphics[width=0.98\textwidth]{figures/figure1.png}
     \caption{
     \textbf{Our in-silico model predicts tracer transport in PVS, SAS and parenchyma following intrathecal injection.}\\
     A) Illustration of the spatial domains of the model: CSF-filled space (ventricles and SAS) $\Omega_{\rm CSF}$, parenchyma $\Omega_{\rm PAR}$ and the PVS surrounding arteries $\Lambda_{\rm artery}$ and the veins $\Lambda_{\rm vein}$ and their interfaces and boundaries (see Methods); 
     B) 3D rendering of the actual geometry used in this study parenchyma, CSF (clipped, blue), arterial network (red), venous network (blue)); 
     C) illustration of the modelled clinical tracer injection procedure; 
     D) simulated tracer concentration after 1, 6, 12, and 24 hours (opacity increases linearly from 0 to $2\,$mol/l); 
     E) sagittal, coronal and axial view of computed tracer concentrations over T1-image after 1, 6, 12 and 24\,h (low concentrations transparent, pial surface in pink, arachnoid membrane in cyan, and arteries in dark red); 
     F) total amount of tracer in each cranial compartment over the first 24\,h; 
     G) average tracer concentration in each compartment over the first 24\,h. }
     \label{fig:results1}
\end{figure}
Using previously published multi-modal magnetic resonance imaging
(MRI) data~\cite{hodneland2019new, deistung2009tof,
  schweser2012quantitative, reichenbach2012future,
  deistung2017overview}, we construct a multiscale computational
representation of the human intracranial compartments consisting of
the CSF spaces and brain parenchyma as three-dimensional (3D) domains
and with the PVSs surrounding major surface arteries and veins as
embedded networks of topologically one-dimensional (1D) curves
(\Cref{fig:results1}A--B). We consider a solute concentration field,
varying in space and time, in the 3D domains and in the PVS networks,
and assume that the solute can cross between these compartments
through semi-permeable membranes. As the drivers and modes of
intracranial transport are under substantial
debate\cite{smith2019going, proulx2021cerebrospinal,
  bohr2022glymphatic, hladky2022glymphatic, betsholtz2024advances},
our first target is to establish a baseline model accounting for a
reasonably conservative set of mechanisms and their integrated effect
over a timescale of several minutes to a few days. To this end, we
assume that the solute will (i) diffuse within all compartments, with
diffusivity depending on the effective properties of the relevant
medium\cite{sykova2008diffusion}; (ii) experience significant
dispersive effects due to the pulsatile flow of CSF induced by the
cardiac and respiratory cycles\cite{vinje2019respiratory,
  sharp2019dispersion, ray2021quantitative, troyetsky2021dispersion};
and (iii) be convected by a (small) net flow of CSF resulting from
production in the choroid plexus with CSF efflux across the upper
convexity\cite{hornkjol2022csf} and from the peristaltic pumping
effect of pulse wave pulsations in surface periarterial
spaces~\cite{mestre2018flow, gjerde2023directional}. Mathematically, this model is represented by a
mixed-dimensional system of coupled time-dependent partial
differential equations\cite{masri2024modelling}, which we solve
numerically with high accuracy using a mass-conserving finite element
scheme and the FEniCS finite element software\cite{alnaes2015fenics,
  kuchta2020assembly} (see Methods). The computational framework and
associated software are all openly available~\cite{ZENODO}.

Simulating a glymphatic MRI protocol~\cite{ringstad2017glymphatic,
  ringstad2018brain, eide2024functional}(\Cref{fig:results1}C), we then predict the
spreading of $0.5$ mmol intrathecally injected Gadobutrol after its
appearance at the craniocervical junction
(\Cref{fig:results1}D--E). After one hour, the in-silico
tracer moves upwards in the SAS frontally of the brainstem, and
quickly reaches the supratentorial regions. Here, it spreads both
posterior through the quadrigeminal cistern and the longitudinal
fissure, and anteriorly through the outer SAS, reaching the top of the
cerebral cortex after around 12 hours. After 24 hours, the tracer
covers most of the brain surface (with the exception of some posterior
regions) and has penetrated substantially into the parenchymal
tissue. This pattern is reflected in the mean tracer concentrations in
each compartment, where the CSF space reaches its peak of
$1.4\,$mmol/l after 2 hours, followed by the arterial PVS
concentration peak ($2.9\,$mmol/l after 4 hours)
(\Cref{fig:results1}G). While the mean concentrations in these
compartments drop soon after peaking, the parenchymal tissue
slowly enriches with tracer over the first 24 hours, with a final
value of $0.3$ mmol/l. Overall, about 40\% of the total amount of
tracer remains in the cranium 24 hours post-injection, with the largest
share in the CSF (53\%), followed by the parenchyma (35\%), the
venous PVS (8\%) and the arterial PVS
(5\%) (\Cref{fig:results1}H).
